\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-08-31 papers/methods/paper.tex
More changes

For supernova-independent uncertainty, the selection functions (a skew normal for low-redshift and an error function for high-redshift) are given independent uncertainty of $0.01$ on all parameters (mean and width for the CDF selection function, and mean, width and skewness for the skew normal selection function). Draw from each survey simulation until we have 1000 LowZ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 250 supernovae for the DES three-year spectroscopic analysis. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys, as designed.

@@

We investigate the cosmological bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. {\green Our current model and techniques can quantify the effect of different scatter models, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model. We note that the incorrect summary statistics do not only bias cosmology, but also cause significant bias with the recovery of the standardisation parameters $\alpha$ and $\beta$. Fixing $\alpha$ and $\beta$ values to simulation values when fitting aids in recovering correct cosmology, but does not remove all bias.}

Unfortunately, adding extra fit parameters to allow for shifting observables washes out cosmology, and applying a specific bias correction requires running a fiducial simulation (assuming cosmology, population and scatter model) and binning data, and is difficult to do whilst accounting for correlations with population and scatter model. This is compounded by the fact that bias corrections do not in general improve fits (increase the log posterior), and so are difficult to fit inherently. Colour bias corrections for the {\celeven} scatter model are included in our cosmology model, the strength of which are controlled by a fit parameter, $s_c$, however neither the {\gten} nor {\celeven} simulation fits constrained this parameter, and thus the majority of bias in the {\celeven} fits remain. Works such as \citet{Kessler2017} show that bias corrections can be applied to supernovae datasets that can robustly handle multiple intrinsic scatter models, and future work will center on uniting these methodologies - incorporating better bias corrections that separate intrinsic scatter bias and non-Gaussian summary statistic bias from Malmquist bias, without having to precompute standardisation parameters and populations.

@@

Table \ref{tab:parameter_correlations} lists the fit correlations to our model fit parameters (excluding the LowZ band systematics, and Malmquist bias uncertainty parameters which had negligible correlation), showing (in order) cosmological parameters, standardisation parameters, population width and skewness parameters, intrinsic dispersion parameters, mass-step parameters, population mean parameters, SALT2 model systematics, dust systematic, global HST calibration systematic, peculiar velocity systematic, global redshift systematic and DES band magnitude and wavelength systematics. Figure \ref{fig:bulk_cor} show the full correlations between all non-systematic model parameters. As expected, $w$ bias is introduced primarily through incorrectly determined $\beta$, which itself is tied to the unexplained colour dispersion. Other interesting correlations are shown and discussed in Figure \ref{fig:bulk_cor}, however none have as significant an impact on cosmology as $\beta$. Table \ref{tab:parameter_correlations} does show that $\langle M_B \rangle$ is strongly correlated with $w$, however it is recovered much more accurately than $\beta$. The band systematics for DES filters $g$, $r$ and $i$ also show significant correlation with $w$, highlighting the importance of minimising instrumental uncertainty.

For the sample size of the DES and LowZ supernova samples (of order 350 supernova), the bias from intrinsic scatter models are sub-dominant to the statistical uncertainty, as shown in Figure \ref{fig:bulk_posterior}. {\green For our full systematics model, the bias represents a deviation between $-0.3\sigma$ to $0.3\sigma$ depending on scatter model, and as such we will leave more complicated treatment of them for future work.}

@@

The importance of validating models using high-precision statistics gained by performing fits to hundreds of data realisations cannot be overstated, however this validation is lacking in many earlier BHM models for supernova cosmology. We have validated this model against many realisations of simplistic simulations with well-known and well-defined statistics, and found no cosmological bias. When validating using SNANA simulations, we find evidence of cosmological bias which is traced back to light curve fits reporting biased observables and incorrect covariance. Allowing fully parametrised corrections on observed supernovae summary statistics is found to make cosmology fits too weak, and allowing simulation based corrections to vary in strength is found to give minor reductions in $w$ bias, however the uncertainty on the intrinsic scatter model itself limits the efficacy of the bias corrections. For the data size represented in the DES three-year spectroscopic survey, the determined biases should be sub-dominant to other sources of uncertainty, however this cannot be expected for future analyses with larger datasets. Stricter bias corrections calculated from simulations are required to reduce bias. Ideally, this would include further work on the physical intrinsic dispersion of the Type Ia supernovae population such that we can better characterise this bias.

@@

When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments which have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:app_like}
\end{align}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{align}
P({\rm data} | S , \theta) P(S, \theta) &= P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &= \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&= \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&= \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) }
\end{align}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{align}
where we define the denominator as $w$ for simplicity in future derivations.

@@

Equations \eqref{eq:disc1} and \eqref{eq:disc2} make the assumption that, for our colour distribution, $\mathcal{N}^{\rm Skew}(\mu, \sigma, \alpha)$ is well approximated by $\mathcal{N}(\mu, \sigma)$. We sought to improve on this approximation by adjusting the mean and standard deviation of the approximated normal to match the actual mean and standard deviation of skew normal. With $\delta \equiv \alpha/\sqrt{1+\alpha^2}$, the correct mean and standard deviation are
\begin{align}
\mu_1 &= \mu_0 + \sqrt{\frac{2}{\pi}} \delta \sigma_0 \\
\sigma_1 &= \sigma_0 \sqrt{1 - \frac{2 \delta^2}{\pi}}.
\end{align}
We can then test the approximation $\mathcal{N}^{\rm Skew}(\mu_0, \sigma_0, \alpha) \rightarrow \mathcal{N}(\mu_1, \sigma_1)$. Unfortunately, this shift to the mean and standard deviation of the normal approximation did not produce stable posterior surfaces which Stan could fit when we included full covariance between $m_B$, $x_1$ and $c$ in the underlying population, and so we simplified the model to have independent parameters. Stable surfaces with underlying population covariance were found when we fixed $\sigma_c$ in the shift correction, such that $\mu_1 = \mu_0 + \sqrt{2/\pi}\delta k$, where we set $k=0.1$ to mirror the width of the input simulation population.  However, this resulted in several population parameters becoming biased, and so we do not fix $\sigma_c$. Comparing whether we shift our normal in the approximation or simply discard skewness, Figure \ref{fig:shift} shows that the calculated efficiency is significantly discrepant to the actual efficiency if the normal approximation is not shifted. The biases when using shifted or unshifted normal approximations when we fit our model on Gaussian and skewed underlying populations are shown in Figure \ref{fig:simple_w_super}, and only the shifted normal approximation correctly recovers underlying population parameters.

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{simple_w_shift_dist_0.pdf}
	\end{center}
	\caption{Marginalised probability distributions for 100 realisations of cosmology, fit to Flat $w$CDM with prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, each containing 1000 simulated high-z and 1000 simulated low-z supernovae. The dashed green surfaces represent a fit to an underlying Gaussian colour population with the unshifted model. The blue solid surface represents fits to a skewed colour population with the unshifted model, and the purple dotted surface represents a fit to a skewed colour population with the shifted model. The superscript $0$ and $1$ denote the two different surveys (high-z and low-z respectively), and similarly the first four $\langle c^i \rangle$ parameters represent the four redshift nodes in the high-z survey, and the last four represent the nodes for the low-z survey. We can see that the shifted model is far better able to recover skewed input populations than the unshifted, performing better in terms of recovering skewness $\alpha_c$, mean colour $\langle c \rangle$ and width of the colour distribution $\sigma_c$. The unshifted model recovers the correct colour mean and width if you approximate a skew normal as a normal: $\Delta\mu = \sqrt{2/\pi}\sigma_c\delta_c \approx 0.071$, which is approximately the deviation found in fits to the colour population mean. Importantly, the unshifted model when run on skewed data (the solid blue) shows extreme bias in $\alpha_c$, where it fits strongly around zero regardless, showing it to be a poor approximation.}
	\label{fig:simple_w_super}
\end{figure*}

@@

The simplest and most commonly found optimisation we employed was to precompute as much as possible to reduce the complexity of the mathematical graph our model is translated into to compute the surface derivatives. For example, when computing the distance modulus, redshift is encountered to various powers. Instead of computing those powers in Stan, we simply pass in several arrays of redshift values already raised to the correct power. Small changes like this however only give small results.

@@

This parametrisation does come with one significant downside - inflexibility. For each step the algorithm takes, we do not recompute the cholesky decomposition of the covariance of the summary statistics - that happens once at the beginning of the model setup. If we had kept the full covariance matrix parametrisation we could modify the matrix easily - for example when incorporating intrinsic dispersion we could simple add on a secondary matrix to create an updated covariance. Using cholesky decompositions, ${\rm cholesky}(\Sigma_1 + \Sigma_2) \neq L_1 + L_2$, and so we would need to recompute the decomposition for each step, which discards most of the computational benefit just gained.

>>>>

For supernova-independent uncertainty, the selection functions (a skew normal for low-redshift and an error function for high-redshift) are given independent uncertainty of $0.01$ on all parameters (mean and width for the CDF selection function, and mean, width and skewness for the skew normal selection function). Draw from each survey simulation until we have 1000 LowZ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 350 supernovae for the combined DES three-year and low-redshift spectroscopic analysis. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys, as designed.

@@

We investigate the cosmological bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty reported from fitting the SALT2 model to light curves is incorrect, non-Gaussian and biased. {\green Our current model and techniques can quantify the effect of different scatter models on biasing the observed summary statistics, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model. We note that the incorrect summary statistics do not only bias cosmology, but also cause significant bias with the recovery of the standardisation parameters $\alpha$ and $\beta$. Fixing $\alpha$ and $\beta$ values to simulation values when fitting aids in recovering correct cosmology, but does not remove all bias.}

Unfortunately, adding extra fit parameters to allow for shifting observables washes out cosmology, and applying a specific bias correction requires running a fiducial simulation (assuming cosmology, population and scatter model) and binning data. It is difficult to do whilst accounting for correlations with population and scatter model. This is compounded by the fact that bias corrections do not in general improve fits (increase the log posterior), and so are difficult to fit inherently. Colour bias corrections for the {\celeven} scatter model are included in our cosmology model, the strengths of which are controlled by a fit parameter, $s_c$, however neither the {\gten} nor {\celeven} simulation fits constrained this parameter, and thus the majority of bias in the {\celeven} fits remain. Works such as \citet{Kessler2017} show that bias corrections can be applied to supernovae datasets that can robustly handle multiple intrinsic scatter models, and future work will center on uniting these methodologies - incorporating better bias corrections that separate intrinsic scatter bias and non-Gaussian summary statistic bias from Malmquist bias, without having to precompute standardisation parameters and populations.

@@

Table \ref{tab:parameter_correlations} lists the fit correlations between our model fit parameters (excluding the LowZ band systematics, and Malmquist bias uncertainty parameters which had negligible correlation), showing (in order) cosmological parameters, standardisation parameters, population width and skewness parameters, intrinsic dispersion parameters, mass-step parameters, population mean parameters, SALT2 model systematics, dust systematic, global HST calibration systematic, peculiar velocity systematic, global redshift systematic and DES band magnitude and wavelength systematics. Figure \ref{fig:bulk_cor} show the full correlations between all non-systematic model parameters. As expected, $w$ bias is introduced primarily through incorrectly determined $\beta$, which itself is tied to the unexplained colour dispersion and biases in colour introduced by the unknown intrinsic scatter model. Other interesting correlations are shown and discussed in Figure \ref{fig:bulk_cor}, however none have as significant an impact on cosmology as $\beta$. Table \ref{tab:parameter_correlations} does show that $\langle M_B \rangle$ is strongly correlated with $w$, however it is recovered much more accurately than $\beta$. The band systematics for DES filters $g$, $r$ and $i$ also show significant correlation with $w$, highlighting the importance of minimising instrumental uncertainty.

For the sample size of the DES and LowZ supernova samples (of order 350 supernova), the bias from intrinsic scatter models is sub-dominant to the statistical uncertainty, as shown in Figure \ref{fig:bulk_posterior}. {\green For our full systematics model, the bias represents a deviation between $-0.3\sigma$ to $0.3\sigma$ depending on scatter model, and as such we will leave more complicated treatment of them for future work.}

@@

The importance of validating models using high-precision statistics gained by performing fits to hundreds of data realisations cannot be overstated, however this validation is lacking in many earlier BHM models for supernova cosmology. We have validated this model against many realisations of simplistic simulations with well-known and well-defined statistics, and found no cosmological bias. When validating using SNANA simulations, we find evidence of cosmological bias which is traced back to light curve fits reporting biased observables and incorrect covariance. Allowing fully parametrised corrections on observed supernovae summary statistics introduces too many degrees of freedom and is found to make cosmology fits too weak. Allowing simulation based corrections to vary in strength is found to give minor reductions in $w$ bias, however the uncertainty on the intrinsic scatter model itself limits the efficacy of the bias corrections. For the data size represented in the DES three-year spectroscopic survey, the determined biases should be sub-dominant to other sources of uncertainty, however this cannot be expected for future analyses with larger datasets. Stricter bias corrections calculated from simulations are required to reduce bias. Ideally, this would include further work on the calculation of intrinsic dispersion of the Type Ia supernovae population such that we can better characterise this bias.

@@

When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:app_like}
\end{align}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{align}
P({\rm data} | S , \theta) P(S, \theta) &= P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &= \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&= \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&= \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) }
\end{align}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{align}
\mathcal{L}&= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{align}
where we define the denominator as $d$ for simplicity in future derivations.

@@

In this section, we investigate the effect of approximating the skew normal underlying colour distribution as a normal. Specifically, equations \eqref{eq:disc1} and \eqref{eq:disc2} make the assumption that, for our colour distribution, $\mathcal{N}^{\rm Skew}(\mu, \sigma, \alpha)$ is well approximated by $\mathcal{N}(\mu, \sigma)$. We sought to improve on this approximation by adjusting the mean and standard deviation of the approximated normal to more accurately describe the actual mean and standard deviation of a skew normal. With $\delta \equiv \alpha/\sqrt{1+\alpha^2}$, the correct mean and standard deviation are
\begin{align}
\mu_1 &= \mu_0 + \sqrt{\frac{2}{\pi}} \delta \sigma_0 \\
\sigma_1 &= \sigma_0 \sqrt{1 - \frac{2 \delta^2}{\pi}},
\end{align}
where we highlight that $\mu$ here represents the mean of the distribution, not distance modulus. We can then test the approximation $\mathcal{N}^{\rm Skew}(\mu_0, \sigma_0, \alpha) \rightarrow \mathcal{N}(\mu_1, \sigma_1)$. Unfortunately, this shift to the mean and standard deviation of the normal approximation where we treat $m_B$, $x_1$ and $c$ as a multivariate skew normal did not produce stable posterior surfaces. Due to this, we treat the underlying $m_B$, $x_1$ and $c$ populations as independent.

We tested a fixed $\sigma_c$ in the shift correction, such that $\mu_1 = \mu_0 + \sqrt{2/\pi}\delta k$, where we set $k=0.1$ to mirror the width of the input simulation population. This resulted in stable posterior surfaces, however this introduced recovery bias in several population parameters, and so we do not fix $\sigma_c$. Comparing whether we shift our normal in the approximation or simply discard skewness, Figure \ref{fig:shift} shows that the calculated efficiency is significantly discrepant to the actual efficiency if the normal approximation is not shifted. The biases when using shifted or unshifted normal approximations when we fit our model on Gaussian and skewed underlying populations are shown in Figure \ref{fig:simple_w_super}, and only the shifted normal approximation correctly recovers underlying population parameters.

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{simple_w_shift_dist_0.pdf}
	\end{center}
	\caption{Marginalised probability distributions for 100 realisations of cosmology, fit to Flat $w$CDM with prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, each containing 1000 simulated high-z and 1000 simulated low-z supernovae. The dashed green surfaces represent a fit to an underlying Gaussian colour population with the unshifted model. The blue solid surface represents fits to a skewed colour population with the unshifted model, and the purple dotted surface represents a fit to a skewed colour population with the shifted model. The superscript $0$ and $1$ denote the two different surveys (high-z and low-z respectively), and similarly the first four $\langle c^i \rangle$ parameters represent the four redshift nodes in the high-z survey, and the last four represent the nodes for the low-z survey. We can see that the shifted model is far better able to recover skewed input populations than the unshifted, performing better in terms of recovering skewness $\alpha_c$, mean colour $\langle c \rangle$ and width of the colour distribution $\sigma_c$. The unshifted model recovers the correct colour mean and width if you approximate a skew normal as a normal: $\Delta\mu = \sqrt{2/\pi}\sigma_c\delta_c \approx 0.071$, which is approximately the deviation found in fits to the colour population mean. Importantly, the unshifted model when run on skewed data (the solid blue) shows extreme bias in $\alpha_c$, where it fits strongly around zero regardless, showing it to be a poor approximation. Based on these results and the fantastic performance in correctly recovering underlying populations of the shifted normal approximation, we adopt that in our model.}
	\label{fig:simple_w_super}
\end{figure*}

@@

The simplest and most commonly found optimisation we employed was to precompute as much as possible to reduce the complexity of the mathematical graph our model is translated into to compute the surface derivatives. For example, when computing the distance modulus, redshift is encountered to various powers. Instead of computing those powers in Stan, we simply pass in several arrays of redshift values already raised to the correct power. Small changes like this however only give small improvements.

@@

This parametrisation does come with one significant downside - inflexibility. For each step the algorithm takes, we do not recompute the cholesky decomposition of the covariance of the summary statistics - that happens once at the beginning of the model setup. If we had kept the full covariance matrix parametrisation we could modify the matrix easily - for example when incorporating intrinsic dispersion we could simple add on a secondary matrix to create an updated covariance. However as the cholesky decomposition of a sum of matrices is not equal to the sum of the cholesky decomposition of each individual matrix, we would need to recompute the decomposition for each step, which discards most of the computational benefit just gained.
